% {\it Maybe better in the Contributions section ?}

The contribution of the thesis author to other published works during the thesis period is listed below.
\bigskip

\noindent Contribution in another project from the Canadian Epilepsy Network (CENet):

\begin{itemize}  
\item F.~F. Hamdan, C.~T. Myers, P.~Cossette, P.~Lemay, D.~Spiegelman, A.~D. Laporte,
  C.~Nassif, O.~Diallo, {\bf J.~Monlong}, M.~Cadieux-Dion, S.~Dobrzeniecka,
  C.~Meloche, K.~Retterer, M.~T. Cho, J.~A. Rosenfeld, W.~Bi, C.~Massicotte,
  M.~Miguet, L.~Brunga, B.~M. Regan, K.~Mo, C.~Tam, A.~Schneider,
  G.~Hollingsworth, D.~R. FitzPatrick, A.~Donaldson, N.~Canham, E.~Blair,
  B.~Kerr, A.~E. Fry, R.~H. Thomas, J.~Shelagh, J.~A. Hurst, H.~Brittain,
  M.~Blyth, R.~R. Lebel, E.~H. Gerkes, L.~Davis-Keppen, Q.~Stein, W.~K. Chung,
  S.~J. Dorison, P.~J. Benke, E.~Fassi, N.~Corsten-Janssen, E.-J. Kamsteeg,
  F.~T. Mau-Them, A.-L. Bruel, A.~Verloes, K.~{\~{O}}unap, M.~H. Wojcik, D.~V.
  Albert, S.~Venkateswaran, T.~Ware, D.~Jones, Y.-C. Liu, S.~S. Mohammad,
  P.~Bizargity, C.~A. Bacino, V.~Leuzzi, S.~Martinelli, B.~Dallapiccola,
  M.~Tartaglia, L.~Blumkin, K.~J. Wierenga, G.~Purcarin, J.~J. O'Byrne,
  S.~Stockler, A.~Lehman, B.~Keren, M.-C. Nougues, C.~Mignot, S.~Auvin,
  C.~Nava, S.~M. Hiatt, M.~Bebin, Y.~Shao, F.~Scaglia, S.~R. Lalani, R.~E.
  Frye, I.~T. Jarjour, S.~Jacques, R.-M. Boucher, E.~Riou, M.~Srour,
  L.~Carmant, A.~Lortie, P.~Major, P.~Diadori, F.~Dubeau, G.~D'Anjou,
  G.~Bourque, S.~F. Berkovic, L.~G. Sadleir, P.~M. Campeau, Z.~Kibar, R.~G.
  Lafreni{\`{e}}re, S.~L. Girard, S.~Mercimek-Mahmutoglu, C.~Boelman, G.~A.
  Rouleau, I.~E. Scheffer, H.~C. Mefford, D.~M. Andrade, E.~Rossignol, B.~A.
  Minassian, and J.~L. Michaud.
\newblock {High Rate of Recurrent De Novo Mutations in Developmental and
  Epileptic Encephalopathies}.
\newblock \emph{The American Journal of Human Genetics}, 101\penalty0
  (5):\penalty0 664--685, nov 2017.
\newblock \doi{10.1016/j.ajhg.2017.09.008}.
\end{itemize}

\noindent Contributions to projects not directly relevant for the thesis topic:

\begin{itemize}
\item M.~Mele, P.~G. Ferreira, F.~Reverter, D.~S. DeLuca, {\bf J.~Monlong}, M.~Sammeth,
  T.~R. Young, J.~M. Goldmann, D.~D. Pervouchine, T.~J. Sullivan, R.~Johnson,
  A.~V. Segre, S.~Djebali, A.~Niarchou, GTEx~Consortium, F.~A. Wright,
  T.~Lappalainen, M.~Calvo, G.~Getz, E.~T. Dermitzakis, K.~G. Ardlie, and
  R.~Guigo.
\newblock {The human transcriptome across tissues and individuals}.
\newblock \emph{Science}, 348\penalty0 (6235):\penalty0 660--665, may 2015.
\newblock \doi{10.1126/science.aaa0355}.

\item D.~D. Pervouchine, S.~Djebali, A.~Breschi, C.~A. Davis, P.~P. Barja, A.~Dobin,
  A.~Tanzer, J.~Lagarde, C.~Zaleski, L.-H. See, M.~Fastuca, J.~Drenkow,
  H.~Wang, G.~Bussotti, B.~Pei, S.~Balasubramanian, {\bf J.~Monlong}, A.~Harmanci,
  M.~Gerstein, M.~A. Beer, C.~Notredame, R.~Guig{\'{o}}, and T.~R. Gingeras.
\newblock {Enhanced transcriptome maps from multiple mouse tissues reveal
  evolutionary constraint in gene expression.}
\newblock \emph{Nature communications}, 6:\penalty0 5903, jan
  2015{\natexlab{a}}.
\newblock \doi{10.1038/ncomms6903}.

\item {\bf J.~Monlong}, M.~Calvo, P.~G. Ferreira, and R.~Guig{\'{o}}.
\newblock {Identification of genetic variants associated with alternative
  splicing using sQTLseekeR}.
\newblock \emph{Nature Communications}, 5\penalty0 (May):\penalty0 4698, aug
  2014.
\newblock \doi{10.1038/ncomms5698}.

\item P.~G. Ferreira, P.~Jares, D.~Rico, G.~Gomez-Lopez, A.~Martinez-Trillos,
  N.~Villamor, S.~Ecker, A.~Gonzalez-Perez, D.~G. Knowles, {\bf J.~Monlong},
  R.~Johnson, V.~Quesada, S.~Djebali, P.~Papasaikas, M.~Lopez-Guerra,
  D.~Colomer, C.~Royo, M.~Cazorla, M.~Pinyol, G.~Clot, M.~Aymerich, M.~Rozman,
  M.~Kulis, D.~Tamborero, A.~Gouin, J.~Blanc, M.~Gut, I.~Gut, X.~S. Puente,
  D.~G. Pisano, J.~I. Martin-Subero, N.~Lopez-Bigas, A.~Lopez-Guillermo,
  A.~Valencia, C.~Lopez-Otin, E.~Campo, and R.~Guigo.
\newblock {Transcriptome characterization by RNA sequencing identifies a major
  molecular and clinical subdivision in chronic lymphocytic leukemia}.
\newblock \emph{Genome Research}, 24\penalty0 (2):\penalty0 212--226, feb 2014.
\newblock \doi{10.1101/gr.152132.112}.

\item T.~Lappalainen, M.~Sammeth*, M.~R. Friedl{\"{a}}nder*, P.~a.~C. {‘t Hoen}*,
  {\bf J.~Monlong}*, M.~a. Rivas*, M.~Gonz{\`{a}}lez-Porta, N.~Kurbatova, T.~Griebel,
  P.~G. Ferreira, M.~Barann, T.~Wieland, L.~Greger, M.~van Iterson,
  J.~Alml{\"{o}}f, P.~Ribeca, I.~Pulyakhina, D.~Esser, T.~Giger, A.~Tikhonov,
  M.~Sultan, G.~Bertier, D.~G. MacArthur, M.~Lek, E.~Lizano, H.~P.~J. Buermans,
  I.~Padioleau, T.~Schwarzmayr, O.~Karlberg, H.~Ongen, H.~Kilpinen, S.~Beltran,
  M.~Gut, K.~Kahlem, V.~Amstislavskiy, O.~Stegle, M.~Pirinen, S.~B. Montgomery,
  P.~Donnelly, M.~I. McCarthy, P.~Flicek, T.~M. Strom, H.~Lehrach,
  S.~Schreiber, R.~Sudbrak, {\'{A}}.~Carracedo, S.~E. Antonarakis,
  R.~H{\"{a}}sler, A.-C. Syv{\"{a}}nen, G.-J. van Ommen, A.~Brazma,
  T.~Meitinger, P.~Rosenstiel, R.~Guig{\'{o}}, I.~G. Gut, X.~Estivill, and
  E.~T. Dermitzakis.
\newblock {Transcriptome and genome sequencing uncovers functional variation in
  humans}.
\newblock \emph{Nature}, sep 2013.
\newblock \doi{10.1038/nature12531}.
\end{itemize}



%%% Local Variables:
%%% mode: latex
%%% TeX-master: "../main"
%%% End:
